Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems

### 2024-06-14 oup-authoring-template.tex
Into OUP template

\documentclass{article}
\usepackage[a4paper, margin=3cm]{geometry}

\usepackage{amsmath}
\usepackage[svgnames]{xcolor}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{tcolorbox}

\usepackage{fancyvrb}
\newenvironment{CodeInput}
{\begin{tcolorbox}[title=input,boxrule=0pt]}
{\end{tcolorbox}}
\newenvironment{CodeOutput}
{\begin{tcolorbox}[title=output,boxrule=0pt]}
{\end{tcolorbox}}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.13,1.00}{#1}}
\newcommand{\ClassTok}[1]{\textcolor[rgb]{0.27,0.56,0.65}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
\newcommand{\ValueTok}[1]{\textcolor[rgb]{0.13,0.57,0.41}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.47,0.37,0.15}{#1}}
\newcommand{\IDETok}[1]{\textcolor[rgb]{0.58,0.58,0.58}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.70,0.27,0.27}{#1}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph

\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

\hypersetup{
  pdftitle={Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
}

@@

\usepackage{authblk}
\title{Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.}
\author[1,2]{Mauro Silberberg}
\author[3]{Henning Hermjakob}
\author[3]{Rahuman S. Malik-Sheriff}
\author[1,2]{Hernán E. Grecco}
\affil[1]{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física. Buenos Aires, Argentina.}
\affil[2]{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA). Buenos Aires, Argentina}
\affil[3]{European Bioinformatics Institute, European Molecular Biology Laboratory (EMBL-EBI), Wellcome Genome Campus, Cambridge, UK}

@@

@@

\begin{abstract}
\Acfp{CRN} play a pivotal role in diverse fields
such as systems biology, biochemistry, chemical engineering, and epidemiology.
High-level definitions of \acp{CRN} enables to use various simulation approaches,
including deterministic and stochastic methods,
from the same model.
However,
existing Python tools for simulation of \ac{CRN} typically wrap external C/C++ libraries for
model definition, translation into equations and/or numerically solving them,
limiting their extensibility and integration with the broader Python ecosystem.
In response, we developed Poincaré and SimBio,
two novel Python packages for simulation of dynamical systems and \acp{CRN}.
Poincaré serves as a foundation for dynamical systems modelling,
while SimBio extends this functionality to \acp{CRN},
including support for the \ac{SBML}.
Poincaré and SimBio are developed as pure Python packages
enabling users to easily extend their simulation capabilities
by writing new or leveraging other Python packages.
Moreover, this does not compromise the performance,
as code can be \ac{JIT} compiled with Numba.
Our benchmark tests using curated models from the BioModels repository
demonstrate that these tools may provide a potentially
superior performance advantage compared to other existing tools.
Additionally, to ensure a user-friendly experience,
our packages use standard typed modern Python syntax that provides a seamless integration with \acp{IDE}.
Our Python-centric approach significantly enhances code analysis, error detection, and refactoring capabilities,
positioning Poincaré and SimBio as valuable tools for the modelling community.
\end{abstract}

\acresetall % reset acronyms after abstract

\hypertarget{introduction}{
  \section{Introduction}
  \label{introduction}
}

@@

\hypertarget{results}{
  \section{Results}
  \label{results}
}

@@

\hypertarget{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}{
  \subsection{Versatile modelling and simulation of dynamical systems with Poincaré}
  \label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}
}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The variable name to which a component is assigned can be automatically saved in the component for introspection
  (i.e., \texttt{Oscillator.x.name\ ==\ "x"}),
\item
  It provides a namespace that allows to easily define multiple independent models in the same script,
\item
  It allows \acp{IDE} to provide autocomplete and refactoring capabilities
  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x}, \texttt{v} and \texttt{eq}),
\item
  It allows creation of instances that can be composed into a bigger model (Figure~\ref{fig-composition}).
\end{enumerate}

@@

\hypertarget{extensible-definition-of-reaction-networks-using-simbio}{
  \subsection{Extensible definition of reaction networks usingSimBio}
  \label{extensible-definition-of-reaction-networks-using-simbio}
}

@@

\hypertarget{reproducibility-and-performance}{
  \subsection{Reproducibility and performance}
  \label{reproducibility-and-performance}
}

@@

\hypertarget{discussion}{
  \section{Discussion}
  \label{discussion}
}

@@

\hypertarget{resources}{
  \section{Resources}
  \label{resources}
}

@@

\bibliographystyle{plain}
\bibliography{bibliography.bib}

\newpage

\input{latex/figure1.tex}
\input{latex/figure2.tex}
\input{latex/figure3.tex}
\input{latex/figure4.tex}
\input{latex/figure5.tex}
\input{latex/figure6.tex}

>>>>

%%
%% Copyright 2022 OXFORD UNIVERSITY PRESS
%%
%% This file is part of the 'oup-authoring-template Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'oup-authoring-template Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
%% with bibliographic references
%%

\documentclass[namedate,numsec,webpdf,modern,large]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}

\graphicspath{{Fig/}}

% line numbers
%\usepackage[mathlines, switch]{lineno}
%\usepackage[right]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

@@

\usepackage{tcolorbox}
\newenvironment{CodeInput}{\begin{tcolorbox}[title=input,boxrule=0pt]}{\end{tcolorbox}}
\newenvironment{CodeOutput}{\begin{tcolorbox}[title=output,boxrule=0pt]}{\end{tcolorbox}}

\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.13,1.00}{#1}}
\newcommand{\ClassTok}[1]{\textcolor[rgb]{0.27,0.56,0.65}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
\newcommand{\ValueTok}[1]{\textcolor[rgb]{0.13,0.57,0.41}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.47,0.37,0.15}{#1}}
\newcommand{\IDETok}[1]{\textcolor[rgb]{0.58,0.58,0.58}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.70,0.27,0.27}{#1}}

@@

\journaltitle{Journal Title Here}
\DOI{DOI HERE}
\copyrightyear{2022}
\pubyear{2019}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems]{Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems}

\author[1,2]{Mauro Silberberg\ORCID{0000-0000-0000-0000}}
\author[3]{Henning Hermjakob\ORCID{0000-0000-0000-0000}}
\author[3]{Rahuman S. Malik-Sheriff\ORCID{0000-0000-0000-0000}}
\author[1,2,$\ast$]{Hernán E. Grecco\ORCID{0000-0000-0000-0000}}

\authormark{Author Name et al.}

\address[1]{\orgname{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
\address[2]{\orgname{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA)}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
\address[3]{\orgdiv{European Bioinformatics Institute}, \orgname{European Molecular Biology Laboratory (EMBL-EBI)}, \orgaddress{\street{Wellcome Genome Campus}, \state{Cambridge}, \country{UK}}}

\corresp[$\ast$]{Corresponding author. \href{email:hgrecco@df.uba.ar}{hgrecco@df.uba.ar}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
%online.}

\abstract{
  \Acfp{CRN} play a pivotal role in diverse fields
  such as systems biology, biochemistry, chemical engineering, and epidemiology.
  High-level definitions of \acp{CRN} enables to use various simulation approaches,
  including deterministic and stochastic methods,
  from the same model.
  However,
  existing Python tools for simulation of \ac{CRN} typically wrap external C/C++ libraries for
  model definition, translation into equations and/or numerically solving them,
  limiting their extensibility and integration with the broader Python ecosystem.
  In response, we developed Poincaré and SimBio,
  two novel Python packages for simulation of dynamical systems and \acp{CRN}.
  Poincaré serves as a foundation for dynamical systems modelling,
  while SimBio extends this functionality to \acp{CRN},
  including support for the \ac{SBML}.
  Poincaré and SimBio are developed as pure Python packages
  enabling users to easily extend their simulation capabilities
  by writing new or leveraging other Python packages.
  Moreover, this does not compromise the performance,
  as code can be \ac{JIT} compiled with Numba.
  Our benchmark tests using curated models from the BioModels repository
  demonstrate that these tools may provide a potentially
  superior performance advantage compared to other existing tools.
  Additionally, to ensure a user-friendly experience,
  our packages use standard typed modern Python syntax that provides a seamless integration with \acp{IDE}.
  Our Python-centric approach significantly enhances code analysis, error detection, and refactoring capabilities,
  positioning Poincaré and SimBio as valuable tools for the modelling community.
}
\keywords{poincare, SimBio, Python, ODE}
\acresetall % reset acronyms after abstract

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

@@

\section{Introduction}

@@

\section{Results}\label{results}

@@

\subsection{Versatile modelling and simulation of dynamical systems with Poincaré}
\label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}

@@

\begin{enumerate}
\item
  The variable name to which a component is assigned can be automatically saved in the component for introspection
  (i.e., \texttt{Oscillator.x.name\ ==\ "x"}),
\item
  It provides a namespace that allows to easily define multiple independent models in the same script,
\item
  It allows \acp{IDE} to provide autocomplete and refactoring capabilities
  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x}, \texttt{v} and \texttt{eq}),
\item
  It allows creation of instances that can be composed into a bigger model (Figure~\ref{fig-composition}).
\end{enumerate}

@@

\subsection{Extensible definition of reaction networks usingSimBio}
\label{extensible-definition-of-reaction-networks-using-simbio}

@@

\subsection{Reproducibility and performance}
\label{reproducibility-and-performance}

@@

\section{Discussion}\label{discussion}

@@

\section{Resources}\label{resources}

@@

%%%%%%%%%%%%%%

\section{Competing interests}
No competing interest is declared.

\section{Author contributions statement}

Must include all authors, identified by initials, for example:
S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.

\section{Acknowledgments}
The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).

% bibliography
\bibliographystyle{abbrvnat}
\bibliography{bibliography}

%% sample for biography with author's image
% \begin{biography}{{\color{black!20}\rule{77pt}{77pt}}}{\author{Author Name.} This is sample author biography text. The values provided in the optional argument are meant for sample purposes. There is no need to include the width and height of an image in the optional argument for live articles. This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
% \end{biography}

%% sample for biography without author's image
% \begin{biography}{}{\author{Author Name.} This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
% \end{biography}

% FIGURE 1
\begin{figure*}[t]

  \begin{minipage}[t]{\linewidth}

  {\centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Derivative}, \ClassTok{System}, \ClassTok{Variable}, \FunctionTok{initial}
  \end{Highlighting}
  \end{CodeInput}

  }

  \end{minipage}%
  \newline
  \begin{minipage}[t]{\linewidth}

  {\centering

  \begin{minipage}[c]{0.6\linewidth}

  {\centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \KeywordTok{class} \ClassTok{Decay}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
      \VariableTok{x}\OperatorTok{:} \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
      \VariableTok{eq} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
  \end{Highlighting}
  \end{CodeInput}

  }

  \end{minipage}%
  %
  \begin{minipage}[c]{0.2\linewidth}

  {\centering

  \[
  \frac{dx}{dt} = -x
  \]

  }

  \end{minipage}%
  %
  \begin{minipage}[c]{0.2\linewidth}

  {\centering

  \[
  x(0) = 1
  \]

  }

  \end{minipage}%

  % \subcaption{\label{fig-first-order}First-order system of an exponential decay. The variable
  % \texttt{x} stores the initial condition for \(x\), and the variable
  % \texttt{eq} stores the rate equation for \(x\).}

  }

  \end{minipage}%
  \newline
  \begin{minipage}[t]{\linewidth}

  {\centering

  \begin{minipage}[c]{0.6\linewidth}

  {\centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \KeywordTok{class} \ClassTok{Oscillator}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
      \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
      \VariableTok{v}: \ClassTok{Derivative} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{(\VariableTok{initial}\OperatorTok{=}\ValueTok{0})}
      \VariableTok{eq} \OperatorTok{=} \VariableTok{v}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
  \end{Highlighting}
  \end{CodeInput}

  }

  \end{minipage}%
  %
  \begin{minipage}[c]{0.2\linewidth}

  {\centering

  \[
  \frac{d^2x}{dt^2} = -x
  \]

  }

  \end{minipage}%
  %
  \begin{minipage}[c]{0.2\linewidth}

  {\centering

  \[
  \begin{cases}
      x(0) &= 1 \\
      \frac{dx}{dt}(0) &= 0
  \end{cases}
  \]

  }

  \end{minipage}%

  % \subcaption{\label{fig-second-order}Second-order system of an harmonic oscillator, where the
  % variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
  % equation is specified for the derivative \texttt{v}.}

  }

  \end{minipage}%
  \newline
  \begin{minipage}[t]{\linewidth}

  {\centering

  \begin{minipage}[c]{0.6\linewidth}

  {\centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \KeywordTok{class} \ClassTok{BigModel}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}:
      \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
      \VariableTok{linked} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\VariableTok{x})}
      \VariableTok{independent} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\ValueTok{2})}
  \end{Highlighting}
  \end{CodeInput}

  }

  \end{minipage}%
  %
  \begin{minipage}[c]{0.2\linewidth}

  {\centering

  \[
  \begin{cases}
      \frac{dx}{dt} = -x \\
      \frac{dx_{ind}}{dt} = -x_{ind}
  \end{cases}
  \]

  }

  \end{minipage}%
  %
  \begin{minipage}[c]{0.2\linewidth}

  {\centering

  \[
  \begin{cases}
      x(0) &= 1 \\
      x_{ind}(0) &= 2
  \end{cases}
  \]

  }

  \end{minipage}%

  % \subcaption{\label{fig-composition}First-order system of two exponential decays by composition of
  % the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
  % to the outer variable \texttt{x}, while the subsystem
  % \texttt{independent} defines a new variable with initial condition $2$,
  % which on the corresponding mathematical expression was named \(x_{ind}\).}

  }

  \end{minipage}%

  \caption{\label{fig-poincare}Code and corresponding mathematical
  expressions for different systems.}

  \end{figure*}

% FIGURE 2
\begin{figure*}[t]

  \begin{minipage}[b]{0.50\linewidth}
  {
      \centering 
      \includegraphics[width=\textwidth]{src/ide/ide1.pdf}
  }
  \end{minipage}%
  %
  \begin{minipage}[b]{0.50\linewidth}
  {
      \centering
      \includegraphics[width=\textwidth]{src/ide/ide2.pdf}
  }
  \end{minipage}%

  \caption{
      \label{fig-ide}
      Screenshots of Visual Studio Code showing
      tooltips (green arrows) and
      highlighted type errors (red arrows).
      On the left,
      we show that \texttt{a},
      a \texttt{Constant} assigned with \texttt{assign(...,\ constant=True)},
      can be used for \texttt{Variable b}'s initial condition.
      Instead,
      it is flagged as a type error (red underlining)
      when using \texttt{c}, a \texttt{Parameter},
      for \texttt{Variable d}'s initial condition,
      The IDE automatically recognizes \texttt{e} as an \texttt{Equation},
      and provides autocompletion of the \texttt{Model}'s components.
      A tooltip is shown when composing models (green arrow, right),
      which show the expected variables and their default values.
      The IDE highlights wrong names (\texttt{z} is not a name in \texttt{Model})
      and mismatched types (\texttt{x} is \texttt{Variable} and \texttt{a} must be a number or a \texttt{Constant})
  }

  \end{figure*}

% FIGURE 3
\begin{figure*}[t]

  \begin{minipage}[t]{\linewidth}

  {\centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \KeywordTok{import} \ClassTok{numpy} \KeywordTok{as} \ClassTok{np}
  \KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Simulator}

  \VariableTok{sim} \OperatorTok{=} \ClassTok{Simulator}\KeywordTok{(\ClassTok{Oscillator})}
  \VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}\KeywordTok{(}\ValueTok{0}, \ValueTok{5}, \ValueTok{101}\KeywordTok{))}
  \end{Highlighting}
  \end{CodeInput}

  }

  \end{minipage}%
  \newline
  \begin{minipage}[t]{0.45\linewidth}

  {\centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \VariableTok{df}.\FunctionTok{head}\KeywordTok{()}
  \end{Highlighting}
  \end{CodeInput}

  \begin{CodeOutput}
      \begin{tabular}[]{@{}lll@{}}
          \toprule
          & x & v \\
          time & & \\
          \midrule
          0.00 & 1.000000 & 0.000000 \\
          0.05 & 0.998750 & -0.049979 \\
          0.10 & 0.995005 & -0.099835 \\
          0.15 & 0.988772 & -0.149440 \\
          0.20 & 0.980069 & -0.198672 \\
          \botrule
      \end{tabular}
  \end{CodeOutput}

  }

  \end{minipage}%
  \hfill
  \begin{minipage}[t]{0.45\linewidth}

  {\centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \VariableTok{df}.\FunctionTok{plot}\KeywordTok{()}\OperatorTok{;}
  \end{Highlighting}
  \end{CodeInput}

  \begin{CodeOutput}
    \includegraphics[width=\textwidth]{article_files/cell-9-output-1}
  \end{CodeOutput}

  }

  \end{minipage}%

  \caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
  from Figure~\ref{fig-second-order}. The output is a
  \texttt{pandas.DataFrame} with a column for each variable and the time
  as index. It is inspected and plotted with the \texttt{pandas} methods.}

  \end{figure*}

% FIGURE 4
\begin{figure*}[t]

  \begin{minipage}[t]{\linewidth}
    \centering 
    \begin{CodeInput}
    \begin{Highlighting}[]
    \KeywordTok{from}\ClassTok{ simbio }\KeywordTok{import}\ClassTok{ Compartment, MassAction, Species, RateLaw, \FunctionTok{initial}}
    \end{Highlighting}
    \end{CodeInput}
  \end{minipage}%
  \newline
  \begin{minipage}[t]{\linewidth}
    \centering 
    \begin{minipage}[c]{0.60\linewidth}
      \centering 
      \begin{CodeInput}
      \begin{Highlighting}[]
        \KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
            \CommentTok{"""2A {-}\textgreater{} B"""}

        \VariableTok{    A}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
        \VariableTok{    B}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
        \VariableTok{    r }\OperatorTok{=}\ClassTok{ RateLaw}\KeywordTok{(}
        \VariableTok{        reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*}\VariableTok{ A}\KeywordTok{]},
        \VariableTok{        products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
        \VariableTok{        rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
        \KeywordTok{    )}
      \end{Highlighting}
      \end{CodeInput}
    \end{minipage}%
    %
    \begin{minipage}[c]{0.40\linewidth}
      \centering 
      \[
      \begin{cases}
          \frac{dA}{dt} = -2 \\
          \frac{dB}{dt} = +1
      \end{cases}
      \begin{cases}
          A(0) &= 1 \\
          B(0) &= 0
      \end{cases}
      \]
    \end{minipage}%
  \end{minipage}%
  \newline
  \begin{minipage}[t]{\linewidth}
      \centering 
      \begin{minipage}[c]{0.60\linewidth}
          \centering 
          \begin{CodeInput}
          \begin{Highlighting}[]
          \KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
              \CommentTok{"""2A {-}\textgreater{} B"""}

          \VariableTok{    A}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
          \VariableTok{    B}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
          \VariableTok{    r }\OperatorTok{=}\ClassTok{ MassAction}\KeywordTok{(}
          \VariableTok{        reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*}\VariableTok{ A}\KeywordTok{]},
          \VariableTok{        products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
          \VariableTok{        rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
          \KeywordTok{    )}
          \end{Highlighting}
          \end{CodeInput}
      \end{minipage}%
      %
      \begin{minipage}[c]{0.40\linewidth}
          \centering 
          \[
          \begin{cases}
              \frac{dA}{dt} = -2 A^2 \\
              \frac{dB}{dt} = +1 A^2
          \end{cases}
          \begin{cases}
              A(0) &= 1 \\
              B(0) &= 0
          \end{cases}
          \]
      \end{minipage}%
  \end{minipage}%

  \caption{
    A reaction system for species \(A\) and \(B\)
    with initial conditions \(1\) and \(0\), respectively. A single reaction
    transforming \(2A\) into \(B\) is saved in variable \texttt{r}. The rate
    \(1\) is specified directly for \texttt{RateLaw}, and is proportional to
    the reactants for \texttt{MassAction}.
  }
  \label{fig-simbio}
\end{figure*}

% FIGURE 5
\begin{figure*}[t]
  \centering

  \begin{CodeInput}
  \begin{Highlighting}[]
  \KeywordTok{from}\ClassTok{ simbio.io }\KeywordTok{import}\ClassTok{ biomodels, sbml}

  \VariableTok{model }\OperatorTok{=}\ClassTok{ sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}  \CommentTok{\# from existing SBML file}
  \VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}  \CommentTok{\# from BioModels}
  \VariableTok{model}
  \end{Highlighting}
  \end{CodeInput}

  \begin{CodeOutput}
      \begin{tabular}{@{}lll@{}}
          \multicolumn{3}{l}{Elowitz2000 - Repressilator} \\
          \toprule
          type & total & names \\
          \midrule
          variables  &  6 & PX, PY, PZ, X, Y, Z \\
          parameters & 17 & cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ... \\
          equations  & 12 & Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ... \\
          \botrule
      \end{tabular}
  \end{CodeOutput}

  \caption{Creation of a model from a local SBML file or one uploaded to BioModels.}
  \label{fig-simbio-io}
\end{figure*}

% FIGURE 6
\begin{figure*}[t]
  \centering
  \includegraphics[width=\textwidth]{src/performance/figures/performance}
  \caption{
    Performance of different softwares to solve models from the curated section of BioModels.
    (left) Run time for the model BIOMD3 as a function of the number of output points.
    (right) Run time for different models for 300 output points,
    using the geometric mean of the different softwares to order them.
    Each point corresponds to the median of 20 runs,
    with a neglibible errorbar given by the interquartile range.
  }
  \label{fig-runtime}
\end{figure*}
